Union and intersection contracts are hard, actually

### 2021-07-29 paper.tex
Make citations uniform

In order to give a precise definition to what values ought to satisfy
union and intersection contracts,
Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
give a coinductively defined semantics inspired by union and
intersection type systems.
The key innovation of their work is recognizing that giving a
semantics to higher-order contracts requires defining not only what
values satisfy a contract, but also what \emph{contexts} satisfy the
contract. This models the situation where context may violate a
contract by calling a function with an inappropriate argument.

Concretely, given a contract $C$,
Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
\rrbracket^-$ of values and contexts, respectively, satisfying the
contract. They are defined by mutual induction and coinduction.

@@

All of this still holds true of the coinductive semantics described in
Section~\ref{sec:coinductive-sem}.
Nevertheless, Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
give an algorithmic system which
is complete for their coinductive semantics. In a remarkable technical
tour de force, their algorithmic system is compatible with
user-defined contracts (see Section~\ref{sec:flat-and-inter}).

A key aspect of the approach of~\cite{KeilThiemannUnionIntersection}
is to rewrite nested union and intersection contracts into a
disjunctive normal form using the De Morgan's law \nickel{A @& (B @|
  C)}$\simeq$\nickel{(A @& B) @| (A @& C)}. The goal is to be able to
delay the choice of branch in intersections as much as possible.

To implement contract verification,
Keil and Thiemann~\cite{KeilThiemannUnionIntersection} resort to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. This aspect is critiqued
in~\cite{RootCauseOfBlame}:
``the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.''

@@

The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on balance, a
technically impressive realization of the coinductive semantics that supports
user-defined contracts, though it is fairly complex and probably difficult to
implement efficiently.

@@

Another realization of the coinductive semantics described in
Section~\ref{sec:coinductive-sem} is given in \cite{RootCauseOfBlame}, which
aims at simplifying the algorithmic system proposed
by~\cite{KeilThiemannUnionIntersection} and described in
Section~\ref{sec:keil-thiemann}.

A key ingredient of~\cite{RootCauseOfBlame} is to disallow user-defined
contracts. This choice gives the authors more freedom in the quest of
a more uniform operational semantics. This is sensible trade-off in
the context of gradual typing à la TypeScript: the problem is to match
contracts with static types, and user-defined contracts don't have a
static type equivalent. On the other hand, the cost would probably not
be worth it for a configuration language like Nickel.

@@

The consequence is that Williams et al.~\cite{RootCauseOfBlame} don't
prove their system complete for the coinductive semantics.
It's probably just an
oversight in the proof: we believe their system to be indeed complete
for the coinductive semantics. But it does speak to the intrinsic
complexity of union and intersection contracts: it is remarkably easy
to get details wrong. This difficulty contrasts with the standard
framework of higher-order contracts where satisfaction is much more
straightforward.

@@

Castagna and Lanvin use abstract interpretation to derive the semantics of
unions and intersections. In their own words, "the resulting definitions are
quite technical and barely intuitive but they have the properties we seek for
[\ldots]".  This makes any comparison with the co-inductive semantics of Kiel
and Thiemann rather difficult. Castagna et al.~\cite{gradualCastagna} build on
Castagna and Lanvin to add polymorphism in the picture. However, it is at the
price of restricting the union and intersection part: it is not possible to
assign intersection types to a function anymore.

>>>>

In order to give a precise definition to what values ought to satisfy
union and intersection contracts,
\citeauthor{KeilThiemannUnionIntersection}~\cite{KeilThiemannUnionIntersection}
give a coinductively defined semantics inspired by union and
intersection type systems.
The key innovation of their work is recognizing that giving a
semantics to higher-order contracts requires defining not only what
values satisfy a contract, but also what \emph{contexts} satisfy the
contract. This models the situation where context may violate a
contract by calling a function with an inappropriate argument.

Concretely, given a contract $C$, \citeauthor{KeilThiemannUnionIntersection}
introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
\rrbracket^-$ of values and contexts, respectively, satisfying the contract.
They are defined by mutual induction and coinduction.

@@

All of this still holds true of the coinductive semantics described in
Section~\ref{sec:coinductive-sem}.  Nevertheless,
\citeauthor{KeilThiemannUnionIntersection} give an algorithmic system which is
complete for their coinductive semantics. In a remarkable technical tour de
force, their algorithmic system is compatible with user-defined contracts (see
Section~\ref{sec:flat-and-inter}).

A key aspect of the approach of \citeauthor{KeilThiemannUnionIntersection}
is to rewrite nested union and intersection contracts into a
disjunctive normal form using the De Morgan's law \nickel{A @& (B @|
  C)}$\simeq$\nickel{(A @& B) @| (A @& C)}. The goal is to be able to
delay the choice of branch in intersections as much as possible.

To implement contract verification, \citeauthor{KeilThiemannUnionIntersection}
resort to specific reduction rules for unions and intersections which perform
this rewriting on the fly. This aspect is critiqued in
\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}: ``the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.''

@@

The algorithmic system of \citeauthor{KeilThiemannUnionIntersection}
is, on balance, a technically impressive realization of the coinductive
semantics that supports user-defined contracts, though it is fairly complex and
probably difficult to implement efficiently.

@@

Another realization of the coinductive semantics described in
Section~\ref{sec:coinductive-sem} is given in
\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}, which aims at simplifying
the algorithmic system proposed by Keil and Thiemann and described in
Section~\ref{sec:keil-thiemann}.

A key ingredient of \citeauthor{RootCauseOfBlame} is to disallow user-defined contracts. This
choice gives the authors more freedom in the quest of a more uniform operational
semantics. This is sensible trade-off in the context of gradual typing à la
TypeScript: the problem is to match contracts with static types, and
user-defined contracts don't have a static type equivalent. On the other hand,
the cost would probably not be worth it for a configuration language like
Nickel.

@@

The consequence is that \citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}
don't prove their system complete for the coinductive semantics.  It's probably
just an oversight in the proof: we believe their system to be indeed complete
for the coinductive semantics. But it does speak to the intrinsic complexity of
union and intersection contracts: it is remarkably easy to get details wrong.
This difficulty contrasts with the standard framework of higher-order contracts
where satisfaction is much more straightforward.

@@

Castagna and Lanvin use abstract interpretation to derive the semantics of
unions and intersections. In their own words, "the resulting definitions are
quite technical and barely intuitive but they have the properties we seek for
[\ldots]".  This makes any comparison with the coinductive semantics of
\citeauthor{KeilThiemannUnionIntersection} rather difficult. Castagna et
al.~\cite{gradualCastagna} build on Castagna and Lanvin to add polymorphism in
the picture. However, it is at the price of restricting the union and
intersection part: it is not possible to assign intersection types to a function
anymore.
